Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2022-10-29 analysis/paper_draft3.Rmd
updated figure and text about batson workflow; moved portion of protoytpe discussion to appendix

```{r batsonworkflowfig, fig.cap="Batson workflow", fig.align = 'center', out.width="50%", fig.pos = "h"}

load_path <- here::here("figures","batson_workflow_fig.png")
knitr::include_graphics(load_path)

@@

A _Batson_ challenge typically proceeds in a three-step sequence (Figure \@ref(fig:batsonworkflowfig)). First, the challenger must meet an initial burden (the _prima facie_ case) of producing just enough evidence of strike bias to raise a sufficient initial inference of illegal bias. If so, then at step two, the striking attorney must produce (permissible) reasons for their challenged strikes. Third, the challenger must discredit those proffered reasons and ultimately persuade the trial judge to find that it is more likely than not that the striking attorney acted with illegal bias.

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases [@flowers2019, 2243; Wash. General Rule 37(g)(v); Calif. Code Civil Procedure $\S$ 231.7(d)(3)(G)]. Prior studies have collected historical strike data in past cases and reported the observed difference in strike rates by race or sex. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or sex of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014, 289-297; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2022.

@@

## The Software Prototype

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and the shiny package [@R; @shiny]. Unlike the simulations, we built the app using Rcpp [@Rcpp] and the Metropolis-Hasting algorithm to sample $b$ from the posterior distribution.

To show how lawyers might use this app in real cases, we loaded this app with real strike data from a convenience sample of attorneys who appeared during jury selection in criminal cases in the federal district court for Connecticut during fiscal years 2013 through 2017. To collect this historical strike data, one of us filed a request with the U.S. District Court of Connecticut based on 28 U.S.C. $\S$ 1868.  Thereafter, we received copies of certain jury-selection records associated with twenty-nine criminal cases in that court during this period. These included strike sheets that indicated the identification number of prospective jurors who were struck by peremptory challenge, the order in which they were struck, and which side (prosecutor or defense) struck which juror. Such records also included a tally of answers to juror questionnaires that asked each prospective juror to report their race and gender.

@@

Now, the credible intervals clearly exclude zero. Thus, we have a credible basis to infer bias against racial-minority prospective jurors in how these attorney use their peremptory challenges in the present case.

Next, suppose the same scenario, except now the strike tally indicates the pattern of strikes by these attorneys against female prospective jurors. If so, we select "gender" as the cognizable class, and the density plots update accordingly (Figure \@ref(fig:figapp3)).

@@

Second, if raised, the tool can help the _Batson_ challenger meet its evidentiary burdens at steps one and three of a _Batson_ challenge (Figure \@ref(fig:batsonworkflowfig)). To meets its evidentiary burdens at the first and third steps, the attorney can introduce the posterior distribution of the striking attorney's bias parameter as relevant statistical evidence. In this way, the tool helps detect bias. But it in no way suffices for deciding a _Batson_ challenge. Such statistical evidence alone cannot account for the veracity of the striking attorney's reasons offered at step two. Besides, _Batson_ procedure allows for any type of circumstantial evidence of bias at steps one and three.

At the same time, US law currently affords an avenue for reducing this missing data problem by systematically collecting historical strike data prospectively. A criminal defendant has a Sixth Amendment right to a public trial that presumptively precludes excluding the public from the voir dire of prospective jurors. Similarly, the public has a presumptive First Amendment right to attend jury voir dire in criminal cases, even if prosecutor and defendant agree to exclude the public [@press1984; @presley2010]. Upon exercise of these rights, courtroom observors can attend jury voir dire and collect relevant strike data based on what they observe in real time.  Even if historical strike data is hard to collect retrospectively, courtroom observers can over time build up a dataset for strike data that they would have been less likely to obtain from court records after the fact.

In practice, the anticipated degree of incompatibility should influence which value of $\alpha$ to select. For example, given perfect compatibility ($\lvert b_{curr} - b_{hist} \rvert = 0$), we should let $\alpha = 1$ (equal weight between current and historical data). This will maximize bias detection even when the number of strikes per trial is small. This effect of increased bias detection can be seen in the simulation results depicted in Figures \@ref(fig:figbd9515), \@ref(fig:figbd9510), and \@ref(fig:figbd956) along the positive diagonals of the individual blocks.

More importantly, in cases where lawyers contest the degree of compatibility, those lawyers should interrogate the choice of $\alpha$ by selecting multiple $\alpha$ values and observing how sensitive the posterior distribution of the striking attorney's bias parameter is to the choice of $\alpha$ value. (Our protoype app permits this kind of testing of sensitivity in real time by providing multiple options for $\alpha$ values.) If the inference of bias is sensitive to the choice of $\alpha$ values, all else equal, then worries about compatibility are warranted and the historical strike data, without more, is less helpful in steps one and three of the _Batson_ procedure than it would otherwise would be. One virtue of this approach is that the choice of $\alpha$ explicitly denotes the degree of weight placed on the historical strike data in the analysis.

Insert appendix here.

>>>>

A _Batson_ challenge typically proceeds in a four-step sequence. The attorney must decide whether to challenge the opposing attorney's use of one or more peremptory strikes as based on illegal bias. Second, the challenger must meet an initial burden (the _prima facie_ case) of producing just enough evidence of strike bias to raise a sufficient initial inference of illegal bias. If so, then at the next step, the striking attorney must proffer permissible reasons for their challenged strikes. Finally, the challenger must discredit those proffered reasons and ultimately persuade the trial judge to find that it is more likely than not that the striking attorney acted with illegal bias.

In any case, evidence of such illegal bias may include data on a party's use of peremptory challenges in past cases [@flowers2019, 2243; Wash. General Rule 37(g)(v); Calif. Code Civil Procedure $\S$ 231.7(d)(3)(G)]. Prior studies have collected such historical strike data in past cases and reported the observed difference in strike rates by race or sex. Typically, they test for the probability of observing a non-zero difference in strike rates by the race or sex of the struck prospective jurors, given repeated sampling from a hypothetical population of peremptory strikes with zero such difference [e.g., @Eisenberg2017; @Grosso2012; for discussion, see @Gastwirth2014, 289-297; @Gastwirth2013]. Prior studies have also modeled how much a prospective juror's race affected the odds of being struck. For different modeling approaches using the same historical strike data from Mississippi, see @Craft2018a; @DeCamp2021; and @Dunn2022.

In turn, the inference of bias based in part on historical strike data is relevant to lawyers at three of the four steps of the _Batson_ challenge, i.e., for deciding whether to argue a _Batson_ violation, for establishing a prima facie case, and for ultimately proving that attorney strike bias in the particular case is more likely than not (Figure \@ref(fig:batsonworkflowfig)).

```{r batsonworkflowfig, fig.cap="Batson workflow", fig.align = 'center', out.width="100%", fig.pos = "h"}

load_path <- here::here("figures","batson_workflow_fig2.pdf")
knitr::include_graphics(load_path)

@@

@@

## The Software Prototype {#prototype}

We describe here a prototype software application ("app") that implements the approach described above and that attorneys and others can use in real time to detect bias in the use of peremptory challenges. We built this app with `r R.version$version.string` and the shiny package [@R; @shiny]. Unlike the simulations, we built the app using Rcpp [@Rcpp] and the Metropolis-Hasting algorithm to sample $b$ from the posterior distribution. To show how lawyers might use this app in real cases, we loaded this app with actual strike data from a convenience sample of attorneys who appeared during jury selection in criminal cases in the federal district court for Connecticut during fiscal years 2013 through 2017.

To collect this historical strike data, one of us filed a request with the U.S. District Court of Connecticut based on 28 U.S.C. $\S$ 1868.  Thereafter, we received copies of certain jury-selection records associated with twenty-nine criminal cases in that court during this period. These included strike sheets that indicated the identification number of prospective jurors who were struck by peremptory challenge, the order in which they were struck, and which side (prosecutor or defense) struck which juror. Such records also included a tally of answers to juror questionnaires that asked each prospective juror to report their race and gender.

@@

Now, the credible intervals clearly exclude zero. Thus, we have a credible basis to infer bias against racial-minority prospective jurors in how these attorney use their peremptory challenges in the present case. In the Appendix, we present additional illustrations wherein the app is used to detect gender bias and where we assign less weight to the historical strike data.

Second, if raised, the tool can help the _Batson_ challenger meet the  evidentiary burdens of presenting a prima facie case and of ultimately proving illegal bias (Figure \@ref(fig:batsonworkflowfig)). To meet those burdens, the attorney can introduce the posterior distribution of the striking attorney's bias parameter as relevant statistical evidence. In this way, the tool helps detect bias.

But as with _any_ inference from historical strike data, the approach here in no way suffices for deciding a _Batson_ challenge. Such statistical evidence alone cannot account for the veracity of the striking attorney's proferred reasons. Besides, _Batson_ procedure allows for any type of circumstantial evidence of bias for purposes of the ultimately proving or disproving illegal bias.

US law currently affords a way to ameliorate this missing data problem. The public has a presumptive First Amendment right to attend jury voir dire in criminal cases, even if prosecutor and defendant agree to exclude the public. Similarly, a criminal defendant has a Sixth Amendment right to a public trial that presumptively precludes excluding the public from the voir dire of prospective jurors [@press1984; @presley2010]. If these rights are asserted, any person can often attend jury voir dire and collect relevant strike data based on what they observe in real time.  Thus, though historical strike data is hard to collect retrospectively, such courtroom observers can prospectively build up a dataset for strike data that they would have been less likely to obtain from court records after the fact.

In practice, the anticipated degree of incompatibility should influence which value of $\alpha$ to select. Given perfect compatibility ($\lvert b_{curr} - b_{hist} \rvert = 0$), we should let $\alpha = 1$ (equal weight between current and historical data). This will maximize bias detection even when the number of strikes per trial is small. This effect of increased bias detection can be seen in the simulation results depicted in Figures \@ref(fig:figbd9515), \@ref(fig:figbd9510), and \@ref(fig:figbd956) along the positive diagonals of the individual blocks.

In cases where lawyers contest the degree of compatibility, those lawyers can and should interrogate the choice of $\alpha$ by selecting multiple $\alpha$ values and observing how sensitive the posterior distribution of the striking attorney's bias parameter is to the choice of $\alpha$ value. (Our protoype app permits this kind of testing of sensitivity in real time by providing multiple options for $\alpha$ values.) If the inference of bias is sensitive to the choice of $\alpha$ values, all else equal, then worries about incompatibility are warranted and the historical strike data, without more, is less helpful to proving a _Batson_ violation than it would otherwise would be. In any case, one virtue of this approach is that the choice of $\alpha$ explicitly and transparently denotes the degree of weight placed on the historical strike data in the analysis.

Here, we extend the discussion in section \@ref(prototype) of the paper by presenting additional illustrations of how a lawyer might use the prototype app.

Suppose the same scenario discussed in section \@ref(prototype), except now the strike tally indicates the pattern of strikes by these attorneys against female prospective jurors. If so, we select "gender" as the cognizable class, and the density plots update accordingly (Figure \@ref(fig:figapp3)).

@@
